---
title: "GIS, a transformational approach"
subtitle: "Part 2: got lines?"
description: |
   Third in a series of posts supporting _Geographic Information Analysis_ 
from: markdown+emoji
author: "David O'Sullivan"
toc: true
lightbox:
  match: auto
code-annotations: hover
code-fold: show
filters: 
  - nutshell
categories:
  - geographic information analysis
  - geospatial
  - R
  - tutorial
execute:
  cache: true
freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
date: 09-12-2025
---


In the [previous post](../2025-09-09-gia-chapter-1B-part-1/index.qmd) I explained the genesis of this short series of 'Got \_\_\_\_?' posts. This post comes hard on the heels of the last one because I'd already written some of it before realising that attempting all sixteen transformations in a single post would be too much of a good(?) thing.

Let's gone!^[Warriors.]

```{r}
#| label: imports
#| code-fold: true
#| output: false
library(spatstat)
library(sfdep)
library(sfnetworks)
library(tidygraph)
library(sf)
library(ggplot2)
library(terra)
library(tidyr)
library(dplyr)
library(ggplot2)
library(cols4all)
library(patchwork)

theme_set(theme_void())
set.seed(1)
```

## From Lines...
As before, we need some data. It will make sense to make both some random lines, and also a network.

Random lines are simple enough. Again, we'll have thirty of them, in time-honoured statistical tradition.

```{r}
#| label: make-lines-data
#| fig-cap: 30 random lines
cx <- 1.745e6  # <1>
cy <- 5.425e6

get_random_line <- function(offset_e, offset_n) {
  st_linestring(matrix(
    rnorm(4, sd = 1000) + rep(c(offset_e, offset_n), 2),
    ncol = 2, byrow = TRUE))
}

lines <- mapply(get_random_line,
                offset_e = rep(cx, 30), 
                offset_n = rep(cy, 30), 
                SIMPLIFY = FALSE) |>
  st_sfc() |>
  data.frame() |>
  st_as_sf(crs = 2193)

ggplot() +
  geom_sf(data = lines)
```
1. So that the lines are centred somewhere near Wellington.

Networks are a little trickier, because they necessarily also include points. The code below uses `spatstat::rMatClust` to first make a clustered set of points, then [`sfdep::st_nb_gabriel`](https://sfdep.josiahparry.com/reference/st_nb_delaunay) and [`sfdep::st_as_graph`](https://sfdep.josiahparry.com/reference/st_as_graph) to make a Gabriel graph among those points. The result `nw` is a [`tidygraph` network object](https://tidygraph.data-imaginist.com/), from which we can extract edges (lines) as a `sf` object.

```{r}
#| label: make-network-data
#| fig-cap: A Gabriel graph based random network
pp <- rMatClust(
  kappa = 1e-6,  # <1>
  scale = 400,   # <2>
  mu = 10,       # <3>
  win = owin(c(-2500, 2500), c(-2500, 2500))
)                # <4>

pp_sf <- pp |>
  as.data.frame() |> 
  mutate(x = cx + x, y = cy + y) |>
  st_as_sf(coords = c("x", "y"), crs = 2193) |>
  mutate(ID = row_number())

nw <- pp_sf |> 
  st_as_sfc() |>
  st_as_graph(
    pp_sf |>
      st_as_sfc() |> 
      st_nb_gabriel())

edges <- nw |>
  activate("edges") |>
  st_as_sf() 

ggplot() + 
  geom_sf(data = edges) + 
  geom_sf_label(data = pp_sf, aes(label = ID), size = 3) +
  theme_void()
```
1. Intensity of the Poisson distributed cluster centres.
2. Size of the clusters (in metres).
3. Mean number of events per cluster.
4. This combination of parameters should give us about 250 events on average.

The network is at least somewhat reminiscent of the connections we might see among a collection of small towns.

## ... to points
Our oracular table demands that we make points from lines by intersection. We already know that all the lines in our network intersect at nodes, so it makes more sense to generate the points of intersection of our random arrangement of lines.

```{r}
#| label: line-intersections
#| fig-cap: Points based on intersection of the random lines
intersections <- lines |>
  st_intersection() |>
  filter(st_geometry_type(geometry) == "POINT")

ggplot() + 
  geom_sf(data = lines, colour = "lightgray") +
  geom_sf(data = intersections)
```

This is very straightforward. The process slows down rather dramatically if there are a lot more objects, but with only 30 lines is quick. It's worth noting that I have to filter out all the 'intersections' that are themselves lines, in other words not points. That's because left to its own devices `st_intersection` treats each line as intersecting with itself. If I omit the filter, you can see what I mean: 

```{r}
#| label: self-intersections
lines |> 
  st_intersection()
```
## ... to lines
Shortest paths are next. This was interesting to figure out. The `tidygraph` semantics are quite nice once you get used to them, but at least at first, and for me, a little confusing. 

`tidygraph` wraps `igraph` and applies many of the `igraph` functions that yield subgraphs (such as shortest path) using a `morph` function, which may yield several subgraphs. For example, a connected subgraphs algorithm might find many potential subgraphs. In such cases, to extract a particular graph of interest another function `crystallise` is used. In a case such as shortest path where only one subgraph is expected, we can morph and crystallise in one step with `convert`. 

So here's what that looks like for shortest path between two vertices, selected using the vertex ID labelled map above, for the fact that they involved diagonally traversing most of the network. I make two different shortest paths, one topological based on counting only the number of edges traversed, and one distance-based where the `weights` parameter for shortest path is calculated from the lengths of edges.

```{r}
#| label: shortest-path
#| fig-cap: Shortest path across the random networks based on counting edges (black), and accumulated distance along edges (red).
sp_edges <- nw |>
  convert(to_shortest_path, from = 166, to = 160) |>
  activate("edges") |>
  st_as_sf()

sp_distances <- nw |>
  convert(to_shortest_path, from = 166, to = 160, weights = st_length(edges)) |>
  activate("edges") |>
  st_as_sf()

ggplot() +
  geom_sf(data = edges, colour = "lightgrey") +
  geom_sf(data = sp_edges, colour = "black") +
  geom_sf(data = sp_distances, colour = "red")
```

The two shortest paths produced are quite different, as we would expect, with the distance-based one a much more directly 'diagonal' path across the space.

I've used `igraph` a fair amount (see e.g. [this post](https://computinggeographically.org/chapters/chap6/fig6-12-graph-drawings.html) and [this project](https://dosull.github.io/antarctica/)) but have found its plotting interface confusing. In addition, maintaining spatial coordinates for vertices is difficult at times, and of course `igraph` knows nothing about cartographic projections. The `sfnetworks` and `tidygraph` combination seems a very promising development in this space.

## ... to areas
After the excitement of network analysis, it seems rather pedestrian to go back to boring old buffering, but the table of all geographical knowledge knows what it wants, and buffers it will be.

```{r}
#| label: line-buffers
line_buffers <- edges |>
  st_buffer(100)

ggplot() +
  geom_sf(data = line_buffers, colour = NA) +
  geom_sf(data = edges, lwd = 0.35)
```

Not a compelling chunk of code, but the visual effect of buffering the lines in a network is very pleasing.

It would perhaps be better accomplished in a drawing package, but it's worth knowing you can create single-sided buffers too.

```{r}
#| label: sided-line-buffers
left_buffers <- edges |>
  st_buffer(100, singleSide = TRUE)

right_buffers <- edges |>
  st_buffer(-100, singleSide = TRUE)

ggplot() +
  geom_sf(data = left_buffers, fill = "orange", colour = NA) +
  geom_sf(data = right_buffers, fill = "limegreen", colour = NA) +
  geom_sf(data = edges, lwd = 2, colour = "white")
```

I am, of course, channeling the [flag of Cóte d'Ivoire](https://en.wikipedia.org/wiki/Flag_of_Ivory_Coast) here, and not the [flag of Ireland](https://en.wikipedia.org/wiki/Flag_of_Ireland).

## ... to fields
As in the previous post concerning transformations from points, the distance to nearest object surface is the output here, and we can adopt a similar approach.

```{r}
#| label: nearest-line-field
target_raster <- edges |>
  as("SpatVector") |>
  rast(res = 25)

target_points <- target_raster |>
  as.points() |>
  st_as_sf()

distance_surface <- target_points |>
  mutate(
    distance = st_distance(geometry, edges) |> 
    apply(MARGIN = 1, min)) |>
  as("SpatVector") |>
  rasterize(target_raster, field = "distance")
names(distance_surface) <- "distance"

ggplot() +
  geom_raster(
    data = distance_surface |> as.data.frame(xy = TRUE),
    aes(x = x, y = y, fill = distance)) +
  geom_sf(data = edges) +
  scale_fill_continuous_c4a_seq(palette = "brewer.prgn")
```

As with the points distance surface output the visual effect here is quite pleasing. Results similar to this may be useful when it comes to the area to line transformation based on polygon skeletonisation.^[Spoilers.]