---
title: "tmap vs. ggplot2 for mapping"
subtitle: "I think I prefer..."
description: "For me at least the choice between ggplot2 and tmap is an ongoing question. Here are my latest thoughts on the subject (with code)."
author: "David O'Sullivan"
toc: true
lightbox:
  match: auto
  loop: true
aliases: 
  - "/posts/2024-11-16-tmap-vs-ggplot/tmap4-vs-ggplot2.html"
code-annotations: hover
filters:
  - nutshell
categories:
  - geospatial
  - R
  - qgis
  - tutorial
  - tmap
  - ggplot
execute: 
  cache: true
  freeze: auto
knitr:
  opts_chunk: 
    warning: false
    message: false
    results: false
date: 11-16-2024
---

I've just spent much of a not especially nice Saturday (weather-wise) tidying up some loose ends on [the _Computing Geographically_ website](https://dosull.github.io/computing-geographically/). In particular, I've been updating the R code snippets that make many of the figures. I started out intending to update the [`tmap` v3](https://cran.r-project.org/web/packages/tmap/index.html) code to [v4](https://github.com/r-tmap/tmap), since the latter is now recommended by the developer. It has, for example, been adopted in the latest edition of the excellent [_Geocomputation with R_](https://r.geocompx.org/).

A little way into the process, I realised that since doing last year's [30 Day Map Challenge](https://30daymapchallenge.com/) (see my efforts [here](https://dosull.github.io/30-day-maps-2023/)) I use `ggplot2` a lot more for my everyday mapping work than I do `tmap`. That's in spite of having taught `tmap` for several years, a choice I made because its learning curve is less steep than `ggplot2`'s. Anyway, long story short, migrating the code on my book website from `tmap` v3 to v4 turned into more of a mixed bag: migrating some code to `tmap` v4, and some to `ggplot2`.

In this post I discuss some things to consider if you are choosing which of these two excellent packages&mdash;that's important: they are _both_ excellent packages&mdash;to use in various situations.  

## Same only different
```{r}
#| code-fold: true

library(sf)
library(terra)
library(tmaptools)
library(tmap)
library(htmlwidgets)
library(cols4all)
library(ggplot2)
library(ggspatial)
library(dplyr)
```

What the packages have in common is that they implement the idea of a [_grammar of graphics_](https://vita.had.co.nz/papers/layered-grammar.html) (hence `ggplot2`'s name) where we progressively add layers to a graphic, specifying for each layer how _data attributes_ are scaled to give _values_ of [:_visual variables_](https://en.wikipedia.org/wiki/Visual_variable) (colour, line width, line style, symbol size and so on). There's a graphic (in German) showing this idea [here](https://medium.com/tdebeus/think-about-the-grammar-of-graphics-when-improving-your-graphs-18e3744d8d18).

So the packages are similar at a high-level. However, I'm not going to delve into details of the differences between these packages. For that, you should explore the extensive online resources available. The best places to start are hard to pick. Maybe [here for `tmap`](https://r.geocompx.org/adv-map) and [here for geospatial stuff in `ggplot2`](https://ggplot2.tidyverse.org/reference/ggsf.html). Just be aware that `tmap` is undergoing a major upheaval as it transitions to v4 and while older code should still work, you'll see a lot of warnings. Meanwhile, `ggplot2` has a habit of changing things without preserving backwards compatibility, so it's advisable to be wary of any code snippets more than 3 or 4 years old when you are looking for help.

Rather than the details, I want to explore the packages 'in use' by looking at four broad aspects of mapping that speak to the advantages of a package specifically designed for mapping (`tmap`) over a more general purpose visualization tool (`ggplot2`), which nevertheless holds it own. Those four things are [choropleth maps](#choropleth-maps-tmaps-killer-app), [raster data](#raster-data-tmaps-other-killer-app), [web maps](#web-maps-tmaps-oth...-ok-this-is-just-getting-silly), and ['map junk'](#map-junk-whatever) (north arrows and the like).

### Choropleth maps: `tmap`'s killer app
Making 'proper' choropleth maps in `ggplot2` is no fun at all. See my [Day 13 experience in the 2023 30 Day Map Challenge](https://dosull.github.io/30-day-maps-2023/maps/13-choropleth.html#ggplot2). The difficulty is that the central tenet of classified choropleth mapping is controlling how you relate data to colour fills. It's not that `ggplot2` won't allow you fine control over this aspect of your choropleth map, it's just that it _really_, _really_ wants you to map your data linearly, and continuously to a colour ramp. That's a reasonable design decision for a general scientific visualization tool. It's just not what cartographers do in choropleth mapping.

To illustrate, here's an old dataset I often use: TB cases in Auckland City in 2006 by Census Area Unit:

```{r}
ak <- st_read("ak-tb.gpkg")
```

And here's the most basic `ggplot2` choropleth map of the `TB_RATE` variable (which is cases per 100,000 population):

```{r}
ggplot() +
  geom_sf(data = ak, aes(fill = TB_RATE)) # <1>
```
1. `aes(fill = ...)` specifies which variable maps to the colour fill

A few things stand out here (to my eye at least):

+ you get a graticule in latitude-longitude even if the data are in a projected coordinate system (to be clear, the map is in the projected coordinates);
+ the default colour ramp goes from dark for low values to light for high values; and
+ the default colour ramp is black-to-blue, which might be preferable to the default muddy browns we've become accustomed to, but is an unusual choice for a map.

Of course we can fix all these things:

```{r}
ggplot() +
  geom_sf(data = ak, aes(fill = TB_RATE)) +
  scale_fill_continuous_c4a_seq( # <1>
    palette = "brewer.reds") +   # <2> 
  coord_sf(datum = st_crs(ak))   # <3>
```
1. `scale_fill_continuous_c4a_seq()` is in the [`cols4all` package](https://github.com/mtennekes/cols4all) which includes a very wide range of palettes
2. Use Brewer palette 'Reds'
3. Get the coordinate system of the data using `st_crs()` and apply to the coordinate frame using `coord_sf()`

Note that I am using a `cols4all` scale, because this is the preferred colour palettes package for `tmap`, and provides a wide array of options. For casual mapping, we likely don't care about the grid, and we can get rid of that too using `theme_void()`:

```{r}
ggplot() +
  geom_sf(data = ak, aes(fill = TB_RATE)) +
  scale_fill_continuous_c4a_seq(
    palette = "brewer.reds") +
  theme_void()
```

Something like the above, which is three lines of code after the introductory `ggplot()` call, is what I use most of the time. And the map is good enough, especially if it's temporary and a stepping stone on the way to some other analysis.

What does the same map look like in `tmap` v4?

```{r}
tm_shape(ak) +
  tm_polygons(
    fill = "TB_RATE",            # <1>
    fill.scale = tm_scale_continuous(
      values = "brewer.reds")) + # <2>
  tm_layout(
    frame = FALSE, 
    legend.frame = FALSE)        # <3>
```
1. Variable name for fill in quotes
2. Brewer 'Reds' again
3. Remove annoying frames around map and legend

Essentially the same map, up to minor aesthetic details. These are easily tweaked in either package, so we won't worry about them too much here. 

Where `tmap` wins out is if you want to experiment with classic choropleth map classification schemes, for example:

```{r}
m1 <- tm_shape(ak) +
  tm_polygons(
    fill = "TB_RATE",
    fill.scale = tm_scale_intervals(values = "brewer.reds")) +
  tm_layout(title = "Pretty", title.size = 0.8,             # <1>
            frame = FALSE, legend.frame = FALSE)

m2 <- tm_shape(ak) +
  tm_polygons(
    fill = "TB_RATE",
    fill.scale = tm_scale_intervals(
      values = "brewer.reds", style = "equal", n = 6)) +    # <2>
  tm_layout(
    title = "Equal intervals", title.size = 0.8,            # <3>
    frame = FALSE, legend.frame = FALSE) 

m3 <- tm_shape(ak) +
  tm_polygons(
    fill = "TB_RATE",
    fill.scale = tm_scale_intervals(
      values = "brewer.reds", style = "quantile", n = 6)) + # <4>
  tm_layout(
    title = "Quantiles", title.size = 0.8,
    frame = FALSE, legend.frame = FALSE)

m4 <- tm_shape(ak) +
  tm_polygons(
    fill = "TB_RATE",
    fill.scale = tm_scale_intervals(
      values = "brewer.reds", style = "sd")) +              # <5>
  tm_layout(title = "Std. Dev.", title.size = 0.8,
            frame = FALSE, legend.frame = FALSE)

tmap_arrange(m1, m2, m3, m4, ncol = 2)
```
1. The default classification style is `"pretty"`, i.e. human-friendly round numbers
2. `style = "equal"` to get equal intervals
3. add a title to show the classification style
4. `style = "quantile"` for quantiles
5. `style = "sd"` for standard deviation breaks

Forget the minor issue with aligning these maps exactly, the point here is that these are all _classified_ choropleth maps, with the classification method specified by the `style` parameter (more options are available than shown here). This is an established way to make choropleth colours easier to parse, or put another way, to make it easier to highlight the specific features of the data you want readers to focus on. 

The magic here is that behind the scenes `tmap` is using the [`classInt` package](https://github.com/r-spatial/classInt/), and if we want to make similar maps in `ggplot2` we have to do that work ourselves: 

```{r}
library(classInt)

class_breaks <- ak$TB_RATE |>
  classIntervals(6, "quantile", digits = 1)
brks <- round(class_breaks$brks, 1)

ggplot() + 
  geom_sf(data = ak, aes(fill = TB_RATE)) +
  scale_fill_binned_c4a_seq(                  # <1>
    palette = "brewer.reds", breaks = brks) + 
  theme_void()
```
1. Note the change to a binned scale, controlled by the breaks

The above is the minimal `ggplot2` version of a quantile map I can come up with. If you want the nice class labels then you have to make those labels yourself and do more work on the legend. You can see an example of this in my [Day 13 map in the 2023 30 Day Map Challenge](https://dosull.github.io/30-day-maps-2023/maps/13-choropleth.html#ggplot2).

Overall, if you are making a lot of classified choropleth maps then `tmap` is your friend.

### Raster data: `tmap`'s other killer app
`tmap` is comfortable dealing with raster data. Here's a simple example:

```{r}
dem <- rast("dem.tif")

tm_shape(dem) + 
  tm_raster(      # <1>
    col = "dem",
    col.scale = tm_scale_intervals(values = "hcl.terrain2"),
    col.legend = tm_legend(title = "Elevation")) +
  tm_layout(legend.frame = FALSE)
```
1. `tm_raster()` 'announces' that these data are raster so that some visual variables might be interpreted differently, e.g. `col` is now what `fill` is on a polygon layer, where `col` is interpreted as linework or outline colour

Nice! Note that the same `tm_scale_intervals()` function is used here to specify colours in this case as it was to specify fill colours in the choropleth map case. We can also have a continuous colour ramp, and layer on top a hillshade. First make the hillshade using some `terra` functions:

```{r}
slope <- dem |> terrain("slope", unit = "radians")
aspect <- dem |> terrain("aspect", unit = "radians")
hillshade <- shade(slope, aspect, 
                   angle = 35, direction = 135)
```

Then just add it as another layer also using the `col` aesthetic with a semi-transparent grey palette.

```{r}
tm_shape(dem) + 
  tm_raster(
    col = "dem",
    col.scale = tm_scale_continuous(values = "hcl.terrain2"),
    col.legend = tm_legend(title = "Elevation")) +
  tm_shape(hillshade) +
  tm_raster(
    col = "hillshade", # <1>
    col.scale = tm_scale_continuous(values = "brewer.greys"),
    col_alpha = 0.35
  ) +
  tm_layout(
    legend.frame = FALSE, legend.show = FALSE)
```
1. `tmap` is OK about adding a second layer using the `col` visual variable

Nicer! If I have one issue here it's with using the parameter name `values` for the palette name, which takes a little bit of getting used to. Of course, the rationale for this is that the palette specifies where the colour visual variable is to get its _values_ which in this context, are colours.

`ggplot2` doesn't really deal in rasters. We _can_ make similar maps pretty easily, by converting to an `x`, `y`, `attribute` dataframe:

```{r}
dem_df <- dem |> as.data.frame(xy = TRUE)
ggplot(dem_df) +
  geom_raster(aes(x = x, y = y, fill = dem)) +             # <1>
  scale_fill_continuous_c4a_seq(
    palette = "hcl.terrain2", name = "Elevation") +
  coord_sf(expand = FALSE) +                               # <2>
  theme_void() +
  theme(
    panel.border = element_rect(fill = NA, linewidth = 1)) # <3>
```
1. Note that for `ggplot2` the colours are still a `fill` here
2. `expand = FALSE` prevents a margin around the data
3. `ggplot2`'s theming system has a lot going on...

You can probably tell I've done a fair bit of this kind of thing: I didn't just magic that `expand = FALSE` option and the `panel.border` thing out of nowhere.

Anyway, while this approach is fine for a small raster like this one, even here the `x-y` dataframe equivalent to the raster has 35,000 or so rows, so this won't scale well. 

And we haven't even got into the headaches involved if you want to layer something else on top of a raster dataset where you need to use the `fill` aesthetic again, like say... a hillshade. Let's just say the [`ggnewscale` package](https://github.com/eliocamp/ggnewscale) is your friend. I leave that as an exercise for the reader...

Again... I think it's clear that `tmap` is a better choice if you are making a lot of maps of raster data.

### Web maps: `tmap`'s oth... OK, this is just getting silly
When it comes to web maps `ggplot2` doesn't even try. `tmap` on the other hand has its 'mode switch' option. Set the mode to `view` and you are making web maps!

```{r results=TRUE}
tmap_mode("view") # <1>
tm_shape(ak) +
  tm_polygons(
    fill = "TB_RATE", 
    fill.scale = tm_scale_intervals(values = "brewer.reds"), 
    fill_alpha = 0.5)
```
1. Now you're making web maps! Use `tmap_mode("plot")` to switch back

&nbsp;

So, yeah, if you are making _bona fide_ web maps, `tmap` every time. Switch back to `plot` mode, and `tmap` also does a good job of allowing you to use a web map as a basemap layer:

```{r}
tmap_mode("plot")
tm_shape(ak) +
  tm_polygons(
    fill = "TB_RATE",
    fill.scale = tm_scale_intervals(values = "brewer.reds"),
    fill_alpha = 0.5) +
  tm_basemap(server = "OpenStreetMap") + # <1>
  tm_layout(frame = FALSE, legend.frame = FALSE)
```
1. Adds a static web map as a basemap

`ggplot2` can get the same effect using the `ggspatial::annotation_maptile()` function (you will also need to have the [`prettymapr` package](http://paleolimbot.github.io/prettymapr/) installed).

```{r}
ggplot() + 
  annotation_map_tile(zoomin = 1) + # <1>
  geom_sf(data = ak, aes(fill = TB_RATE), alpha = 0.5) +
  scale_fill_continuous_c4a_seq(palette = "brewer.reds") +
  theme_void()
```
1. `annotation_map_tile()` is a function from `ggspatial`

Overall though, another win for `tmap`.

### Map junk: whatever
As you might guess from the title of this section, I am not much concerned with north arrows and scalebars and such-like. `tmap` has them built in. 

```{r}
tm_shape(ak) + 
  tm_polygons(
    fill = "TB_RATE",
    fill.scale = tm_scale_continuous(values = "brewer.reds")) +
  tm_compass() +
  tm_scalebar(position = c("left", "bottom")) +
  tm_layout(legend.frame = FALSE)
```

To get the same things in `ggplot2` you need the [`ggspatial` package](https://paleolimbot.github.io/ggspatial/), and they're perfectly serviceable. I am about as excited about this as I sound.

```{r}
ggplot() +
  geom_sf(data = ak, aes(fill = TB_RATE)) +
  scale_fill_continuous_c4a_seq(palette = "brewer.reds") +
  annotation_north_arrow(location = "br") + # <1>
  annotation_scale(location = "bl") +       # <2>
  theme_void() + 
  theme(panel.border = element_rect(fill = NA))
```
1. Another `ggspatial` function
2. And another

Needless to say, both `tmap` and `ggplot2` offer lots of flexibility for adding titles and subtitles and positioning all these elements around the map wherever you want them.

All in all, this one is probably a tie, assuming you've installed `ggspatial`.

## Final thoughts
In brief then:  `tmap` is better at: 

+ classified choropleth maps;
+ dealing with geospatial rasters; and
+ web maps,

and about equal on 

+ web map derived basemaps; and
+ map junk;

The latter two assuming that you have installed `ggspatial`. The funny thing is, I still find myself using `ggplot2` more! For me the counterpoints to the above are:

+ unclassified choropleths are often fine for a quick look-see at your data;
+ I don't make many raster-based maps, and `ggplot2::geom_tile` is often good enough for my purposes;
+ Where I previously used web maps in `tmap` to get an overview of my data, now I tend to use QGIS. If I _really_ need a web map, I will certainly use `tmap`; and
+ I am not a fan of map junk (hence: junk). Don't get me wrong, I'll put a north arrow and scalebar on a map [if necessary](https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/put-a-north-arrow-on-it/), I just don't make many maps like that.

If your mix of use-cases is different, especially if choropleth maps, raster data, and web maps loom larger in your world than they do in mine, then you might wind up making a different choice.

I think the reason I end up gravitating to `ggplot2` is that it is the entry point into a much wider visualization world. The idiom it has popularised, which `tmap` has reworked for mapping, of layering a series of _aesthetics_ each linking a data variable to a visual variable (that so-called _grammar of graphics_) is more cleanly implemented in `ggplot2`, as you'd expect it to be. And the exact same semantics apply to [scatterplots, boxplots, violin plots, histograms, bar charts, and so on](https://ggplot2.tidyverse.org/reference/#layers). The `ggplot2` ecosystem is sprawling and extremely powerful, and making maps is just one small corner of it.

Because of that, I am always making charts using `ggplot(data) + geom_*()` and it feels very natural to make maps the same way.

Having said that, for R coded maps that cover all the mapping bases, you should probably also get to grips with `tmap`. I use both regularly, even if I do use `ggplot2` a little _more_ regularly!